\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

\title{Stability of ecological and epidemiological models via
representation as Generalized Lotka-Volterra dynamics}

\author{
    Stefano Allesina
    \thanks{stefanoallesina.github.io}
   \\
    Department of Ecology \& Evolution \\
    University of Chicago \\
  Chicago, IL 60637 USA \\
  \texttt{\href{mailto:sallesina@uchicago.edu}{\nolinkurl{sallesina@uchicago.edu}}} \\
   \And
    Zachary R. Miller
   \\
    Department of Ecology \& Evolution \\
    University of Chicago \\
   \\
  \texttt{\href{mailto:zachmiller@uchicago.edu}{\nolinkurl{zachmiller@uchicago.edu}}} \\
  }



% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{fancybox}


\begin{document}
\maketitle

\def\tightlist{}


\begin{abstract}
Enter the text of your abstract here.
\end{abstract}

\keywords{
    Lotka-Volterra
   \and
    Stability
   \and
    Quasi-polynomial
   \and
    Lyapunov function
  }

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Several models for ecological, evolutionary, and epidemiological
dynamics can be recast as a Generalized Lotka-Volterra model via the
so-called \emph{quasi-monomial transformation}. This fact was discovered
at least three times (as far as we can tell, independently) by different
authors in the late '80s (Peschel and Mende 1986).

\hypertarget{generalized-lotka-volterra-model}{%
\section{Generalized Lotka-Volterra
model}\label{generalized-lotka-volterra-model}}

\label{sec:glv}

\begin{equation}
\label{eq:glv}
\dot{x}_i = x_i \left(r_i + \sum_{j=1}^m A_{ij} x_j \right)
\end{equation}

\hypertarget{quasi-polynomial-systems}{%
\section{Quasi-Polynomial systems}\label{quasi-polynomial-systems}}

\label{sec:qp}

We now introduce a generalization of Eq. \ref{eq:glv}, defining the
class of quasi-polynomial systems (QP-systems):

\begin{equation}
\label{eq:qp}
\dot{y}_i = y_i \left( s_i + \sum_{j = 1}^m M_{ij} \prod_{k = 1}^n y_k^{B_{jk}} \right)
\end{equation}

where we have \(n\) equations, \(\dot{y}_1, \ldots, \dot{y}_n\). The
vector \(s\) is of length \(n\), \(M\) is a matrix of size
\(n \times m\) containing real coefficients, and \(B\) a matrix of size
\(m \times n\), also containing real coefficients. If \(n = m\), and
thus both \(M\) and \(B\) are square matrices, and further that
\(B=I_n\) (the identity matrix of size \(n\)), the model reduces to the
Generalized Lotka-Volterra model in Eq. \ref{eq:glv} with \(r = s\) and
\(A = M\). If \(B\) contains only integers, Eq. \ref{eq:qp} defines a
\emph{polynomial} system of differential equations; relaxing this
condition to admit \(B\) composed of real numbers, we obtain a
\emph{quasi-polynomial} (QP-) system.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\textbf{QP-representation of Leslie-Gower predator-prey model}

The Leslie-Gower model is simple variation on the classic Lotka-Volterra
predator-prey model. We have two equations:

\begin{equation}
\label{eq:lg}
\begin{cases}
\dot{y}_1 = y_1 (\rho_1 - y_1- \alpha_1 y_2)\\
\dot{y}_2 = y_2 \left(\rho_2 - \alpha_2 \frac{y_2}{y_1} \right)
\end{cases}
\end{equation}

with \(y_1\) representing the prey, \(y_2\) the predator, and all
coefficients are assumed to be positive. The system differs from GLV in
that we have a ratio between the predator and prey in the equation for
the predator. The system is however in QP form, as seen by defining:

\begin{equation}
\label{eq:lgqp}
s = \begin{pmatrix}
\rho_1\\
\rho_2
\end{pmatrix} \quad 
M = \begin{pmatrix}
-1 & -\alpha_1 & 0\\
0 & 0 & -\alpha_2
\end{pmatrix} \quad
B = \begin{pmatrix}
1 & 0 \\
0 & 1 \\
1 & -1
\end{pmatrix}
\end{equation}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{from-quasi-polynomial-to-generalized-lotka-volterra}{%
\section{From Quasi-Polynomial to Generalized
Lotka-Volterra}\label{from-quasi-polynomial-to-generalized-lotka-volterra}}

\label{sec:qptoglv}

We define a set of \(m\) \emph{quasi-monomials}:

\begin{equation}
\label{eq:quasimono}
x_j = \prod_{k=1}^n y_k^{B_{jk}}
\end{equation}

A simple way to identify quasi-monomials for any system that can be
written in QP form is to consider the per capita dynamics:

\begin{equation}
\dot{\log y}_i = \frac{\dot{y}_i}{y_i} = s_i + \sum_{j = 1}^m M_{ij} \prod_{k = 1}^n y_k^{B_{jk}}
\end{equation}

As such, the set of variables, or product of powers of variables,
appearing in the equations for the equations \(\dot{\log y}_i\) defines
the quasi-monomials \(x\).

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\textbf{Quasi-monomials for the Leslie-Gower model}

For the Leslie-Gower model in Eq. \ref{eq:lg} we identify three
quasi-monomials:

\begin{equation}
\label{eq:lgqm}
\begin{cases}
x_1 = y_1^1 \, y_2^0 = y_1\\
x_2 = y_1^0 \, y_2^1 = y_2\\
x_3 = y_1^{-1} \, y_2^1 = \frac{y_2}{y_1}
\end{cases}
\end{equation}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

Now we show how the \(n-\)dimensional QP-system of differential
equations in Eq. \ref{eq:qp} can be recast as an \(m-\)dimensional GLV
system in Eq. \ref{eq:glv}. By chain rule, we have:

\begin{equation}
\label{eq:quasimonodt}
\begin{aligned}
\dot{x}_j &= \sum_k B_{jk}\, \dot{y}_k \, y_{k}^{(B_{jk} -1)} \prod_{l \neq k} y_{l}^{B_{jl}}\\
&=\sum_k B_{jk}\, \frac{\dot{y}_k}{y_k} \prod_{l} y_{l}^{B_{jl}}\\
&=\sum_k B_{jk}\, \frac{\dot{y}_k}{y_k} x_j\\
&=x_j \sum_k B_{jk}\, \frac{\dot{y}_k}{y_k} \\
&=x_j \left(\sum_k B_{jk} s_k + \sum_k B_{jk} \sum_l M_{kl} \prod_l y_l^{B_{kl}} \right)\\
&=x_j \left(\sum_k B_{jk} s_k + \sum_k B_{jk} \sum_l M_{kl} x_l \right)\\
&=x_j \left((B s)_j + \sum_l (B M)_{jl} x_l \right)\\
&=x_j \left(r_j + \sum_l A_{jl} x_l \right)
\end{aligned}
\end{equation}

where we have defined \(A = BM\) and \(r = B s\); \((Bs)_j\) is the
\(j^{\text{th}}\) element of the vector \(Bs\) and \((B M)_{jl}\) is the
coefficient in row \(j\) and column \(l\) of the matrix \(BM\).

Via the quasi-monomial transformation in Eq. \ref{eq:quasimono} we have
turned the \(n-\)dimensional QP-system in Eq. \ref{eq:qp} into an
\(m-\)dimensional GLV system in Eq. \ref{eq:glv}.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\textbf{GLV representation of the Leslie-Gower model}

We can represent the Leslie-Gower model in Eq. \ref{eq:lg} as a
three-dimensional GLV model defined by the quasi-monomials in Eq.
\ref{eq:lgqm} and:

\begin{equation}
r = B s = \begin{pmatrix}
\rho_1\\
\rho_2\\
\rho_2 - \rho_1
\end{pmatrix}
\quad
A = B M = \begin{pmatrix}
-1 & -\alpha_1 & 0\\
0 & 0 & -\alpha_2\\
1 & \alpha_1 & -\alpha_2
\end{pmatrix}
\end{equation}

Note that \(A\) is rank deficient, given that the third row can be
written as the difference between the second and first row.
Rank-deficiency of \(A\) is expected whenever \(m > n\) (as here, where
we went from two to three equations).

The GLV representation of the model becomes:

\begin{equation}
\label{eq:lgglv}
\begin{cases}
\dot{x}_1 = x_1 (\rho_1 - x_1 - \alpha_1 x_2)\\
\dot{x}_2 = x_2 (\rho_2 - \alpha_2 x_3)\\
\dot{x}_3 = x_3 (\rho_2 - \rho_1 + \alpha_1 x_1 - \alpha_2 x_3)
\end{cases}
\end{equation}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{stability-in-generalized-lotka-volterra}{%
\section{Stability in Generalized
Lotka-Volterra}\label{stability-in-generalized-lotka-volterra}}

\hypertarget{lyapunovs-direct-method}{%
\subsection{Lyapunov's direct method}\label{lyapunovs-direct-method}}

Provided with a dynamical system \(\dot{y}_i\)

\hypertarget{candidate-lyapunov-function-for-generalized-lotka-volterra}{%
\subsection{Candidate Lyapunov function for Generalized-Lotka
Volterra}\label{candidate-lyapunov-function-for-generalized-lotka-volterra}}

Before proceeding, we note that if Eq. \ref{eq:glv} has a feasible
(i.e., positive) coexistence equilibrium, then \(r = -A x^\star\), and
as such we can rewrite the equations more compactly as
\(\dot{x}_i = x_i \sum_{j} A_{ij} (x_j - x_j^\star) = x_i \sum_{j} A_{ij} \Delta x_j\),
where we have defined the deviation from equilibrium
\(\Delta x_j = x_j - x_j^\star\).

We now consider the candidate Lyapunov function proposed by Goh:

\begin{equation}
\label{eq:goh}
\begin{cases}
V_{x_i} = x_i - x_i^\star - x_i^\star \log \frac{x_i}{x_i^\star}\\
V = \sum_i w_i V_{x_i}
\end{cases}
\end{equation}

where each \(V_{x_i} > 0\) for any positive \(x_i \neq x_i^\star\) and
zero at the equilibrium. The weights \(w_1, \ldots w_n\) are
non-negative, and at least one is positive. In such cases, \(V\) is
radially unbounded. If we can find weights such that
\(\dot{V} = \sum_i w_i \dot{V}_{x_i} \leq 0\), we can prove stability
(possibly, through LaSalle's invariance principle). Deriving:

\begin{equation}
\label{eq:gohstab}
\begin{aligned}
\dot{V} =& \sum_i w_i \left(\dot{x}_i - x_i^\star \dot{\log x_i} \right)\\
 =& \sum_i w_i \left(x_i \sum_{j} A_{ij} \Delta x_j - x_i^\star \sum_j A_{ij} \Delta x_j \right)\\
 =& \sum_i w_i \left(\Delta x_i \sum_{j} A_{ij} \Delta x_j \right)\\
 =& \sum_i \Delta x_i w_i \sum_j A_{ij} \Delta x_j\\
 =& \sum_i \sum_j w_i A_{ij} \Delta x_i \Delta x_j
\end{aligned}
\end{equation}

Note that in the sum over \(i\) and \(j\), only the symmetric part of
the matrix \(D(w) A = (w_i A_{ij})\) matters (the skew symmetric part
cancels). It is therefore convenient to define a new, symmetric matrix
\(G =\frac{1}{2} (D(w)A + A^T D(w))\), so that our expression becomes:

\begin{equation}
\label{eq:stabcond}
\dot{V} = \frac{1}{2}\sum_i \sum_j (w_i A_{ij} + A_{ji} w_j) \Delta x_i \Delta x_j = \sum_i \sum_j G_{ij} \Delta x_i \Delta x_j
\end{equation}

A symmetric matrix \(G\) satisfying
\(z^T G z = \sum_i \sum_j G_{ij} z_i z_j < 0\) for every \(z \neq 0\) is
called \emph{negative definite}. If the sum can be zero for some
\(z \neq 0\), \(G\) is called \emph{negative semi-definite}. A
symmetric, negative definite matrix has all eigenvalues real and
negative; in a negative semi-definite matrix eigenvalues can be zero. As
such, if we can identify suitable, positive (nonnegative) weights \(w\)
such that \(G\) is negative (semi-)definite, then \(\dot{V} \leq 0\) and
we can prove the stability of the equilibrium \(x^\star\).

\hypertarget{stability-in-qp-systems}{%
\subsection{Stability in QP-systems}\label{stability-in-qp-systems}}

In the Generalized Lotka-Volterra model in Eq. \ref{eq:glv}, the
variables \(x_i\) can in principle take any positive value (at least as
an initial condition), and therefore each \(\Delta x_i\) is radially
unbounded: \(\Delta x_i \in [-x_i^\star, \infty)\); moreover, we can set
(again, at least initially) each \(x_i\) to any arbitrary value,
irrespective of the value of the rest of the \(x_j\). In such a setting,
it is therefore difficult to prove stability via Eq. \ref{eq:stabcond}
if the matrix \(G\) is not negative (semi-)definite.

Note however that, when we represent an \(n-\)dimensional QP-system
using an \(m-\)dimensional GLV system, the quasi monomials \(x_i\) are
function of the original variables \(y_i\). This in turn means that the
perturbations in the GLV system are a function of the perturbations in
the original system: in particular
\(\Delta x_i = \prod_{k=1}^n y_k^{B_{ik}} - \prod_{k=1}^n (y_k^\star)^{B_{ik}}\).

In practice, this means that not all perturbations \(\Delta x\) are
allowed---rather, only those compatible with the definition of the
quasi-monomials. In turn, this means that we could (and often will) find
nonnegative weights in Eq. \ref{eq:stabcond} such that
\(\dot{V} \leq 0\) and yet the matrix \(G\) \emph{is not negative
semi-definite}. In such cases, \(G\) \emph{acts} like a negative
semi-definite matrix on the \emph{admissible} perturbations, i.e., those
abiding by the form specified by the quasi-monomials.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\textbf{Stability of the Leslie-Gower model}

We consider the candidate Lyapunov function in Eq. \ref{eq:goh} for the
QP-representation of the Leslie-Gower model (Eq. \ref{eq:lgglv}). A
convenient choice of weights is
\(w = (0, \rho_1 \alpha_1 + \alpha_2, \rho_2 \alpha_2 )^T\), yielding:

\begin{equation}
G =\frac{1}{2} (D(w)A + A^T D(w)) =
\begin{pmatrix}
0 & 0 & \frac{\rho_2 \alpha_2}{2} \\
0 & 0 & -\frac{\alpha_2^2}{2} \\
\frac{\rho_2 \alpha_2}{2} & -\frac{\alpha_2^2}{2} & -\rho_2 \alpha_2^2
\end{pmatrix}
\end{equation}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{examples}{%
\section{Examples}\label{examples}}

\hypertarget{a-susceptible-infected-recovered-model-with-demography}{%
\subsection{A susceptible-infected-recovered model with
demography}\label{a-susceptible-infected-recovered-model-with-demography}}

\label{sec:sir}

We consider a simple S-I-R model in which mortality in all classes is
counterbalanced by the birth of susceptible individuals:

\begin{equation}
\label{eq:sir}
\begin{cases}
\begin{aligned}
\dot{y}_1 = \delta - \delta y_1 - \beta y_1 y_2 &= y_1 \left(-\delta + \delta \frac{1}{y_1} - \beta y_2 \right)\\
\dot{y}_2 = - (\delta + \gamma) y_2 + \beta y_1 y_2 &= y_2 \left(-(\delta + \gamma) + \beta y_1 \right)\\
\dot{y_3} = \gamma y_2 - \delta y_3 &= y_3 \left(-\delta + \gamma \frac{y_2}{y_3} \right)
\end{aligned}
\end{cases}
\end{equation}

where \(y_1\) represents the proportion of susceptible individuals in
the population, \(y_2\) that of the infected/infectious individuals, and
\(y_3\) the recovered individuals. The parameter \(\delta\) serves both
as a birth rate for the population, and as the mortality rate in each
compartment, \(\beta\) is the transmission rate, and \(\gamma\) the
recovery rate. We have \(y_1 + y_2 + y_3 = 1\). It is well known that if
the critical threshold
\(\mathcal R_0 = \frac{\beta}{\gamma + \delta} > 1\), the model has a
globally stable endemic equilibrium in which a constant proportion of
individuals is infected.

At the endemic equilibrium, we have \(y_1^\star = 1 / \mathcal R_0\),
\(y_2^\star = (\mathcal R_0 - 1) \delta / (\delta + \gamma)\), and
\(y_2^\star = (\mathcal R_0 - 1) \gamma / (\delta + \gamma)\), such that
the equilibrium is feasible only if \(\mathcal R_0 >1\).

We write the model as a four-dimensional GLV system, and probe the
stability of the endemic equilibrium assuming that it is positive.
First, we identify the quasi-monomials and associated matrices and
vectors:

\begin{equation}
\begin{aligned}
x = \begin{pmatrix}
y_1\\
y_2\\
\frac{1}{y_1}\\
\frac{y_2}{y_3}
\end{pmatrix}
\quad
s = \begin{pmatrix}
-\delta \\
-\delta - \gamma \\
-\delta
\end{pmatrix}
\quad
M = \begin{pmatrix}
0 & -\beta & \delta & 0\\
\beta & 0 & 0 & 0 \\
0 & 0 & 0 & \gamma
\end{pmatrix}
\quad
B = \begin{pmatrix}
1 & 0 & 0\\
0 & 1 & 0\\
-1 & 0 & 0\\
0 & 1 & -1
\end{pmatrix}
\\
r = \begin{pmatrix}
-\delta \\
-\delta - \gamma \\
\delta\\
-\gamma
\end{pmatrix}
\quad 
A = \begin{pmatrix}
0 & -\beta & \delta & 0 \\
\beta & 0 & 0 & 0\\
0 & \beta & -\delta & 0 \\
\beta & 0 & 0 & -\gamma
\end{pmatrix}
\end{aligned}
\end{equation}

Now consider the candidate Lyapunov function with weights
\(w = (1,1,0,0)^T\), and the perturbations
\(\Delta x = (y_1 - y_1^\star, y_2 - y_2^\star, 1/y_1 - 1 / y_1^\star, y_2 / y_3 - y_2^\star / y_3^\star)^T\).
Deriving with respect to time, we obtain:

\begin{equation}
\dot{V} = \frac{1}{2} \Delta x^T (D(w) A + A^T D(w)) \Delta x =  - \delta \frac{(y_1 - y_1^\star)^2}{y_1 y_1^\star} \leq 0
\end{equation}

Note that this choice of weights does not result in a negative
semi-definite matrix: the nonzero eigenvalues of \((D(w) A + A^T D(w))\)
are \(\pm \delta / 2\).

Because \(\dot{V} = 0\) whenever \(y_1 = y_1^\star\), to prove stability
we need to make sure that the equilibrium is the only trajectory
contained in the set of points satisfying \(\dot{V} = 0\).

\textbf{LaSalle's}

\hypertarget{stability-of-a-consumer-resource-model-with-inputs}{%
\subsection{Stability of a consumer-resource model with
inputs}\label{stability-of-a-consumer-resource-model-with-inputs}}

We consider a model with \(2n\) equations, equally divided into
resources (\(z_i\)) and consumers (\(y_i\)). The model can be written
as:

\begin{equation}
\label{eq:cr}
\begin{cases}
\dot{z}_i = \kappa_i - \delta_i z_i - z_i \sum_j C_{ij} y_j\\
\dot{y}_i = - \mu_i y_i + y_i \epsilon_i \sum_j C_{ji} z_j
\end{cases}
\end{equation}

where \(\kappa_i\) is the input to resource \(i\), \(\delta_i\) its
degradation rate, \(C_{ij}\) the consumption of \(i\) by consumer \(j\),
\(\mu_i\) the mortality rate of consumer \(i\), and \(\epsilon_i\) the
efficiency of transformation of resources into consumers. We want to
show that, whenever a feasible equilibrium exists, it is globally
stable.

We identify \(3n\) quasi-monomials: \(x_i = z_i\) for
\(i \in \{1, \ldots, n\}\), \(x_i = y_i\) for
\(i \in \{n + 1, \ldots, 2 n\}\), and finally \(x_i = 1 / z_i\) for
\(i \in \{2 n + 1, \ldots, 3 n\}\). This allows us to rewrite the system
in Eq. \ref{eq:cr} as a QP-system defined by:

\begin{equation}
\label{eq:crqp}
s = \begin{pmatrix}
-\delta \\
-\mu \\
\end{pmatrix} \quad 
M = \begin{pmatrix}
0_{n \times n} & -C & D(\kappa)\\
D(\epsilon)C^T & 0_{n \times n} & 0_{n \times n}
\end{pmatrix} \quad
B = \begin{pmatrix}
I_n & 0_{n \times n}\\
0_{n \times n} & I_n\\
-I_n & 0_{n \times n}
\end{pmatrix}
\end{equation}

where \(0_{n \times n}\) is a matrix of size \(n \times n\) containing
zeros, \(I_n\) is the identity matrix of size \(n\) and \(D(\theta)\)
the diagonal matrix with \(\theta\) on the diagonal. We rewrite the
system as a \(3n-\)dimensional GLV defined by:

\begin{equation}
r = B s = \begin{pmatrix}
-\delta \\
-\mu \\
\delta
\end{pmatrix} \quad
A = B M = \begin{pmatrix}
0_{n \times n} & -C & D(\kappa)\\
D(\epsilon) C^T & 0_{n \times n} & 0_{n \times n}\\
0_{n \times n} & C & -D(\kappa)
\end{pmatrix}
\end{equation}

We now consider the candidate Lyapunov function in Eq. \ref{eq:goh} with
weights \(w = (1_n, 1 / \epsilon, 0_n)\). Our matrix \(G\) becomes:

\begin{equation}
G = \frac{1}{2}(D(w) A + A^T D(w)) = \begin{pmatrix}
0_{n \times n} & 0_{n \times n} & \frac{1}{2} D(\kappa)\\
0_{n \times n} & 0_{n \times n} & 0_{n \times n}\\
\frac{1}{2} D(\kappa) & 0_{n \times n} & 0_{n \times n}
\end{pmatrix}
\end{equation}

The matrix is clearly non negative semi-definite. And yet, when we
consider the perturbations:
\(\Delta x = (z - z^\star, y - y^\star, 1 / z - 1 / z^\star)^T\), the
derivative with respect to time in Eq. \ref{eq:stabcond} reduces to:

\begin{equation}
\dot{V} = \sum_i k_i (z_i - z_i^\star) \left( \frac{1}{z_i} - \frac{1}{z_i^\star} \right) = - \sum_i \frac{k_i}{z_i z_i^\star} (z_i - z_i^\star)^2 \leq 0
\end{equation}

\textbf{TODO LaSalle's}

\hypertarget{reducing-nonlinearities-in-a-model-with-higher-order-interactions}{%
\subsection{Reducing nonlinearities in a model with higher-order
interactions}\label{reducing-nonlinearities-in-a-model-with-higher-order-interactions}}

\label{sec:hoi}

The last few years have seen a resurgence in the interest for so-called
higher-order interactions. When including interactions of more than two
populations at a time, dynamics can be altered dramatically. For
example, Grilli \textit{et al.} showed stabilization for the replicator
equation when three or more players interact. The simplest model is that
for a replicator equation describing a three-player game of
rock-paper-scissors:

\begin{equation}
\label{eq:rpshoi}
\begin{cases}
\dot{y}_1 = y_1 (2 y_3^2 + y_1 y_3 - 2 y_1 y_2 - y_2^2)\\
\dot{y}_2 = y_2 (2 y_1^2 + y_1 y_2 - 2 y_2 y_3 - y_3^2)\\
\dot{y}_3 = y_3 (2 y_2^2 + y_2 y_3 - 2 y_1 y_3 - y_2^2)
\end{cases}
\end{equation}

with dynamics occurring on the simplex \(y_1 + y_2 + y_3 = 1\). The
system has a single feasible equilibrium
\(y_i^\star = \frac{1}{3} \quad \forall i\). Note that the system is in
QP-form as defined by the monomials:

\begin{equation}
\label{eq:rpshoimon}
x = 
\begin{pmatrix}
y_1^2\\ y_2^2 \\ y_3^2 \\ y_1 y_2 \\ y_1 y_3 \\ y_2 y_3
\end{pmatrix} 
\end{equation}

and parameters

\begin{equation}
\label{eq:rpshoiqp}
\lambda = \begin{pmatrix}
0\\
0\\
0
\end{pmatrix} \quad 
M = \begin{pmatrix}
0 & -1 & 2 & -2 & 1 & 0\\
2 & 0 & -1 & 1 & 0 & -2\\
-1 & 2 & 0 & 0 & -2 & 1
\end{pmatrix} \quad
B = \begin{pmatrix}
2 & 0 & 0 \\
0 & 2 & 0 \\
0 & 0 & 2 \\
1 & 1 & 0 \\
1 & 0 & 1 \\
0 & 1 & 1
\end{pmatrix}
\end{equation}

As such, the model in Eq. \ref{eq:rpshoi} with three variables
(equivalent to two equations, given the constrain of dynamics happening
in the simplex) and higher-order interactions can be recast as a GLV
model with six variables, and parameters:

\begin{equation}
r = B \lambda = \vec{0}
\quad
A = B M = \begin{pmatrix}
0 & -2 & 4 & -4 & 2 & 0 \\ 
4 & 0 & -2 & 2 & 0 & -4 \\ 
-2 & 4 & 0 & 0 & -4 & 2 \\ 
2 & -1 & 1 & -1 & 1 & -2 \\ 
-1 & 1 & 2 & -2 & -1 & 1 \\ 
1 & 2 & -1 & 1 & -2 & -1
\end{pmatrix}
\end{equation}

Note that matrix \(A\) has eigenvalues
\(-\frac{3}{2}(1 \pm 3 \sqrt{3})\) and an additional four eigenvalues
equal to zero (i.e., it has rank 2, the number of independent equations
in the original system).

By choosing the candidate Lyapunov in Eq. \ref{eq:goh} function with
weights \(\gamma = (1,1,1, 2,2,2)^T\) and considering the perturbations
\(\Delta x = (y_1^2 - {y_1^\star}^2, y_2^2 - {y_2^\star}^2, y_3^2 - {y_3^\star}^2, y_1 y_2 - y_1^\star y_2^\star, y_1 y_3 - y_1^\star y_3^\star, y_2 y_3 - y_2^\star y_3^\star)^T\),
we obtain:

\begin{equation}
\dot{V} = -\frac{2}{3} \left(y_1^2 + y_2^2 + y_3^2 - y_1 y_2 -y_1 y_3 - y_2 y_3 \right) \leq 0
\end{equation}

\textbf{A good proof of this? It is definitely true}

The derivative is thus negative for any \(0<y_i<1\) but the equilibrium
point, proving the stability of the equilibrium.

You can insert references. Here is some text (Kour and Saabne 2014b,
2014a) and see Hadash et al. (2018).

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-hadash2018estimate}{}%
Hadash, Guy, Einat Kermany, Boaz Carmeli, Ofer Lavi, George Kour, and
Alon Jacovi. 2018. {``Estimate and Replace: A Novel Approach to
Integrating Deep Neural Networks with Existing Applications.''}
\emph{arXiv Preprint arXiv:1804.09028}.

\leavevmode\hypertarget{ref-kour2014fast}{}%
Kour, George, and Raid Saabne. 2014a. {``Fast Classification of
Handwritten on-Line Arabic Characters.''} In \emph{Soft Computing and
Pattern Recognition (SoCPaR), 2014 6th International Conference of},
312--18. IEEE.

\leavevmode\hypertarget{ref-kour2014real}{}%
---------. 2014b. {``Real-Time Segmentation of on-Line Handwritten
Arabic Script.''} In \emph{Frontiers in Handwriting Recognition (ICFHR),
2014 14th International Conference on}, 417--22. IEEE.

\leavevmode\hypertarget{ref-peschel1986predator}{}%
Peschel, Manfred, and Werner Mende. 1986. \emph{The Predator-Prey Model:
Do We Live in a Volterra World?} Springer.

\end{CSLReferences}

\bibliographystyle{unsrt}
\bibliography{references.bib}


\end{document}
